---
title: "Pathway analysis"
author: "AmauryPelzer"
date: "19/06/22"
output:
 md_document:
    variant: markdown_github
always_allow_html: true
---

## Introduction

In this workflow, pathway analysis will be performed on host-metagenomics data.

## R environment setup

First, we need to make sure all required packages are installed

```{r setup, warning=FALSE, message=FALSE}
# Check if libraries are already installed > otherwise install it
if(!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager",repos = "http://cran.us.r-project.org")
if(!"rstudioapi" %in% installed.packages()) BiocManager::install("rstudioapi")
if(!"clusterProfiler" %in% installed.packages()) BiocManager::install("clusterProfiler")
if(!"KEGGREST" %in% installed.packages()) BiocManager::install("KEGGREST")
if(!"org.EcK12.eg.db" %in% installed.packages()) BiocManager::install("org.EcK12.eg.db")
if(!"ggplot2" %in% installed.packages()) BiocManager::install("ggplot2")
if(!"tidyverse" %in% installed.packages()) BiocManager::install("tidyverse")

# Load packages
library(rstudioapi)
library(clusterProfiler)
library(KEGGREST)
library(org.EcK12.eg.db)
library(stringr)
library(ggplot2)
library(tidyverse)

# Set your working environment to the location where your current source file is saved into.
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
# Include some functions adapted from ArrayAnalysis.org scripts
WORK.DIR <- getwd()
```

## Data Preparations

The following section will prepare input data to be used in the analysis

```{r read data,warning=FALSE, message=FALSE}
# Read  the metagenomics data file
mgxCount <- read.table(file = '../Data/ecs_relab_2', sep = '\t', header = TRUE)

# Read metadata file sample labels
sampleLabels <- read.table(file = "../Data/metadata", row.names=1, sep = '\t', stringsAsFactors = TRUE)

# Select the samples IDs
mgxCount <- subset(mgxCount, select=rownames(sampleLabels))

# Add column names to the metadata file
colnames(sampleLabels) <- c( "sampleID", "biopsy_location", "disease")
# Check whether sample names are in same order
all(colnames(mgxCount) == rownames(sampleLabels))

sampleLabels$disease <- relevel(sampleLabels$disease,ref="nonIBD")
```

## Filtering Steps

We will apply some filtering process to filter out genes in the input data

```{r filtering,warning=FALSE, message=FALSE}
# Remove genes which has all zero values for all samples then start DE analysis
mgxCount[is.na(mgxCount)] <- 0
nonzero <- rowSums(mgxCount) > 0
mgxCount %<>% .[nonzero,]
```

## T-tests

```{r}
# Separate CD, UC and nonIBD
cd <- sampleLabels[sampleLabels$disease=="CD",]
uc <- sampleLabels[sampleLabels$disease=="UC",]
nonIBD <- sampleLabels[sampleLabels$disease=="nonIBD",]

# Select metagenomics data from CD, UC and nonIBD
metagenomics_cd <- subset(mgxCount, select= rownames(cd))
metagenomics_uc <- subset(mgxCount, select= rownames(uc))
metagenomics_nonIBD <- subset(mgxCount, select= rownames(nonIBD))
```

```{r}
# Compute statistical significance (using t-test)
pvalue_cd = NULL # Empty list for the p-values
tstat_cd = NULL # Empty list of the t test statistics
regulation_cd = NULL # Empty list for the factor: over-represented and under-represented

for(i in 1 : nrow(metagenomics_nonIBD)) { # For each gene : 
	x = metagenomics_nonIBD[i,] # control of gene number i
	y = metagenomics_cd[i,] # CD of gene number i
	
	# Compute t-test between the two conditions
	t = t.test(x, y)
	
	# Put the current p-value in the pvalues list
	pvalue_cd[i] = t$p.value
	# Put the current t-statistic in the tstats list
	tstat_cd[i] = t$statistic
	
	# Put the status (over- or under-represented) in regulation list
	if (rowMeans(x)<rowMeans(y))
	{
	  regulation_cd[i] = "Over-represented"
	}
	else
	{
	  regulation_cd[i] = "Under-represented"
	}
}

#padjust_cd = p.adjust(pvalue_cd, method="BH", n= nrow(metagenomics_nonIBD))
```

```{r}
# Compute statistical significance (using t-test)
pvalue_uc = NULL # Empty list for the p-values
tstat_uc = NULL # Empty list of the t test statistics
regulation_uc = NULL # Empty list for the factor: over-represented and under-represented

for(i in 1 : nrow(metagenomics_nonIBD)) { # For each gene : 
	x = metagenomics_nonIBD[i,] # control of gene number i
	y = metagenomics_uc[i,] # UC of gene number i
	
	# Compute t-test between the two conditions
	t = t.test(x, y)
	
	# Put the current p-value in the pvalues list
	pvalue_uc[i] = t$p.value
	# Put the current t-statistic in the tstats list
	tstat_uc[i] = t$statistic
	
	# Put the status (over- or under-represented) in regulation list
	if (rowMeans(x)<rowMeans(y))
	{
	  regulation_uc[i] = "Over-represented"
	}
	else
	{
	  regulation_uc[i] = "Under-represented"
	}
}

#padjust_uc = p.adjust(pvalue_uc, method="BH")

```

```{r}
# Select statistically significant enzymes
deg.CD <- metagenomics_cd[which(pvalue_cd<0.05),]
deg.CD$pvalue <- pvalue_cd[which(pvalue_cd<0.05)]
deg.CD$regulation <- regulation_cd[which(pvalue_cd<0.05)]

deg.UC <- metagenomics_uc[which(pvalue_uc<0.05),]
deg.UC$pvalue <- pvalue_uc[which(pvalue_uc<0.05)]
deg.UC$regulation <- regulation_uc[which(pvalue_uc<0.05)]
```

## Species-specific pathway analysis

Codes for KEGG organisms can be found here (<https://www.genome.jp/kegg/catalog/org_list.html>)

```{r}
# Subset Bacteroides xylanisolvens
deg.CD.xyl <- deg.CD[grepl("xylanisolvens", rownames(deg.CD), fixed = TRUE),]
deg.UC.xyl <- deg.UC[grepl("xylanisolvens", rownames(deg.UC), fixed = TRUE),]

# Add row names as a column
deg.CD.xyl$Gene.Family <- rownames(deg.CD.xyl)
deg.UC.xyl$Gene.Family <- rownames(deg.UC.xyl)

# Split Gene.Family and Name.Organism
deg.CD.xyl[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(deg.CD.xyl$Gene.Family, ': ', 2)
deg.UC.xyl[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(deg.UC.xyl$Gene.Family, ': ', 2)
```

```{r}
# Add row names as column
deg.CD$Gene.Family <- rownames(deg.CD)
deg.UC$Gene.Family <- rownames(deg.UC)

# Separate Gene.Family from the Name and Organism
deg.CD[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(deg.CD$Gene.Family, ': ', 2)
deg.UC[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(deg.UC$Gene.Family, ': ', 2)

deg.CD[c('Name', 'Organism')] <- str_split_fixed(deg.CD$Name.Organism, '.s__', 2)
deg.UC[c('Name', 'Organism')] <- str_split_fixed(deg.UC$Name.Organism, '.s__', 2)

# Clean char variables
deg.CD$Name <- sub("\\|.*", "", deg.CD$Name)
deg.UC$Name <- sub("\\|.*", "", deg.UC$Name)

deg.CD$Organism <- sub(" .*", "", deg.CD$Organism)
deg.UC$Organism <- sub(" .*", "", deg.UC$Organism)

deg.CD$Organism <- sub("\t.*", "", deg.CD$Organism)
deg.UC$Organism <- sub("\t.*", "", deg.UC$Organism)
```


### PW analysis for CD

```{r}
# Convert EC numbers to Entrez IDs
gene.CD.xyl <- clusterProfiler::bitr(deg.CD.xyl$Gene.Family, fromType = "ENZYME", toType = "ENTREZID", OrgDb = org.EcK12.eg.db)

# Convert Entrez IDs to KEGG IDs
geneList.CD.xyl <- sub("^", "ncbi-geneid:", gene.CD.xyl[,2])
geneList.CD.xyl <- keggConv("eco", geneList.CD.xyl)

# Remove preceding 'eco:'
geneList.CD.xyl <- gsub("eco:", "", geneList.CD.xyl)

# Perform KEGG pathway over-representation analysis
kk.CD.xyl <- enrichKEGG(gene         = geneList.CD.xyl,
                 organism     = 'eco',
                 pvalueCutoff = 0.05)

# Show results
kk.CD.xyl@result
```

```{r}
# Bar plot for thesis
kk.CD.xyl@result[which(kk.CD.xyl@result$p.adjust < 0.2),] %>% 
  ggplot(aes(reorder(Description, Count), Count)) + 
  geom_col(aes(fill = p.adjust)) + 
  scale_fill_gradient2(low = "#0000ff", 
                       mid = "#ff0000",
                       high= "yellow",
                       midpoint = 0.1) + 
  coord_flip() + 
  labs(x = "Pathways (CD)", y = "Gene count")
```


### PW analysis for UC

```{r}
# Convert EC numbers to Entrez IDs
gene.UC.xyl <- clusterProfiler::bitr(deg.UC.xyl$Gene.Family, fromType = "ENZYME", toType = "ENTREZID", OrgDb = org.EcK12.eg.db)

# Convert Entrez IDs to KEGG IDs
geneList.UC.xyl <- sub("^", "ncbi-geneid:", gene.UC.xyl[,2])
geneList.UC.xyl <- keggConv("eco", geneList.UC.xyl)

# Remove preceding 'eco:'
geneList.UC.xyl <- gsub("eco:", "", geneList.UC.xyl)

kk.UC.xyl <- enrichKEGG(gene         = geneList.UC.xyl,
                 organism     = 'eco',
                 pvalueCutoff = 0.05)

# Show results
kk.UC.xyl@result
```

```{r}
# Bar plot for thesis
kk.UC.xyl@result[which(kk.UC.xyl@result$p.adjust < 0.2),] %>% 
  ggplot(aes(reorder(Description, Count), Count)) + 
  geom_col(aes(fill = p.adjust)) + 
  scale_fill_gradient2(low = "#0000ff", 
                       mid = "#ff0000",
                       high= "yellow",
                       midpoint = 0.1) + 
  coord_flip() + 
  labs(x = "Pathways (UC)", y = "Gene count")
```


## Community level pathway analysis

```{r}
# Read  the pathway abundance data file
mgxPW <- read.table(file = '../Data/Mgx_pathabundance', sep = '\t', header = TRUE)

# Select the samples IDs
mgxPW <- subset(mgxPW, select=rownames(sampleLabels))

# Remove genes which has all zero values for all samples then start DE analysis
mgxPW[is.na(mgxPW)] <- 0
nonzero <- rowSums(mgxPW) > 0
mgxPW %<>% .[nonzero,]

# Select metagenomics data from CD, UC and nonIBD
pathway_cd <- subset(mgxPW, select= rownames(cd))
pathway_uc <- subset(mgxPW, select= rownames(uc))
pathway_nonIBD <- subset(mgxPW, select= rownames(nonIBD))
```

```{r}
# Compute statistical significance (using t-test)
pvalue_PW_cd = NULL # Empty list for the p-values
tstat_PW_cd = NULL # Empty list of the t test statistics
regulation_PW_cd = NULL # Empty list for the factor: over-represented and under-represented
score_PW_cd = NULL

for(i in 1 : nrow(pathway_nonIBD)) { # For each gene : 
	x = pathway_nonIBD[i,] # control of gene number i
	y = pathway_cd[i,] # CD of gene number i
	
	# Compute t-test between the two conditions
	t = t.test(x, y)
	
	# Put the current p-value in the pvalues list
	pvalue_PW_cd[i] = t$p.value
	# Put the current t-statistic in the tstats list
	tstat_PW_cd[i] = t$statistic
	
	score_PW_cd[i] = abs(rowMeans(x)-rowMeans(y))
	
	# Put the status (over- or under-represented) in regulation list
	if (rowMeans(x)<rowMeans(y))
	{
	  regulation_PW_cd[i] = "Over-represented"
	}
	else
	{
	  regulation_PW_cd[i] = "Under-represented"
	}
}

#padjust_PW_cd = p.adjust(pvalue_PW_cd, method="BH")
```

```{r}
# Compute statistical significance (using t-test)
pvalue_PW_uc = NULL # Empty list for the p-values
tstat_PW_uc = NULL # Empty list of the t test statistics
regulation_PW_uc = NULL # Empty list for the factor: over-represented and under-represented
score_PW_uc = NULL

for(i in 1 : nrow(pathway_nonIBD)) { # For each gene : 
	x = pathway_nonIBD[i,] # control of gene number i
	y = pathway_uc[i,] # CD of gene number i
	
	# Compute t-test between the two conditions
	t = t.test(x, y)
	
	# Put the current p-value in the pvalues list
	pvalue_PW_uc[i] = t$p.value
	# Put the current t-statistic in the tstats list
	tstat_PW_uc[i] = t$statistic
	
	score_PW_uc[i] = abs(rowMeans(x)-rowMeans(y))
	
	# Put the status (over-) in regulation list
	if (rowMeans(x)<rowMeans(y))
	{
	  regulation_PW_uc[i] = "Over-represented"
	}
	else
	{
	  regulation_PW_uc[i] = "Under-represented"
	}
}

#padjust_PW_uc = p.adjust(pvalue_PW_uc, method="BH", n= nrow(pathway_nonIBD))
```


```{r}
# Select statistically significant enzymes
dePW.CD <- pathway_cd[which(pvalue_PW_cd<0.05),]
dePW.CD$pvalue <- pvalue_PW_cd[which(pvalue_PW_cd<0.05)]
dePW.CD$regulation <- regulation_PW_cd[which(pvalue_PW_cd<0.05)]
dePW.CD$score <- score_PW_cd[which(pvalue_PW_cd<0.05)]
dePW.UC <- pathway_uc[which(pvalue_PW_uc<0.05),]
dePW.UC$pvalue <- pvalue_PW_uc[which(pvalue_PW_uc<0.05)]
dePW.UC$regulation <- regulation_PW_uc[which(pvalue_PW_uc<0.05)]
dePW.UC$score <- score_PW_uc[which(pvalue_PW_uc<0.05)]
```

```{r}
# Add row names as column
dePW.CD$Gene.Family <- rownames(dePW.CD)
dePW.UC$Gene.Family <- rownames(dePW.UC)

# Separate Gene.Family from the Name and Organism
dePW.CD[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(dePW.CD$Gene.Family, ': ', 2)
dePW.UC[c('Gene.Family', 'Name.Organism')] <- str_split_fixed(dePW.UC$Gene.Family, ': ', 2)

dePW.CD[c('Name', 'Organism')] <- str_split_fixed(dePW.CD$Name.Organism, '.s__', 2)
dePW.UC[c('Name', 'Organism')] <- str_split_fixed(dePW.UC$Name.Organism, '.s__', 2)

# Clean char variables
dePW.CD$Name <- sub("\\|.*", "", dePW.CD$Name)
dePW.UC$Name <- sub("\\|.*", "", dePW.UC$Name)

dePW.CD$Organism <- sub(" .*", "", dePW.CD$Organism)
dePW.UC$Organism <- sub(" .*", "", dePW.UC$Organism)

dePW.CD$Organism <- sub("\t.*", "", dePW.CD$Organism)
dePW.UC$Organism <- sub("\t.*", "", dePW.UC$Organism)
```


```{r}
dePW.UC[which(dePW.UC$pvalue<0.0005),] %>% 
  ggplot(aes(reorder(Name, score), score)) + 
  geom_col(aes(fill = pvalue)) + 
  scale_fill_gradient2(low = "#0000ff", 
                       mid = "#ff0000",
                       high= "yellow",
                       midpoint = 0.001) + 
  coord_flip() + 
  labs(x = "Pathways (UC)", y = "Difference in relative abundance")
```

```{r}
dePW.CD[which(dePW.CD$pvalue<0.0005),] %>% 
  ggplot(aes(reorder(Name, score), score)) + 
  geom_col(aes(fill = pvalue)) + 
  scale_fill_gradient2(low = "#0000ff", 
                       mid = "#ff0000",
                       high= "yellow",
                       midpoint = 0.001) + 
  coord_flip() + 
  labs(x = "Pathways (CD)", y = "Difference in relative abundance")
```
